---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
---


## Loading and preprocessing the data

1. Load the data  

```{r loaddata}
if (!file.exists("activity.csv")){
  unzip(zipfile="activity.zip")
}  
data1 <- read.csv("activity.csv", sep = ",", colClasses = c("numeric", "Date", "numeric"))
```

2. Process/Transform the data

```{r processData}
data1$interval <- as.factor(data1$interval)
```

## What is mean total number of steps taken per day?

Calculate the mean total number of steps taken per day.  
```{r calculateSteps}
dailySteps <- aggregate(steps ~ date, data1, sum,na.rm=TRUE)
```

Create a histogram of the total number of steps taken per day.

```{r histStepsPerDay}
hist(dailySteps$steps, 
     col = "green",
     breaks = seq(from = 0, to = 27000, by = 1500),
     main = "Histogram of Daily Steps", 
     xlab = "Total Daily Steps",
     ylab = "Timer per day")

```

Calculate the mean and median steps taken daily  

```{r}
meanSteps <- mean(dailySteps$steps,na.rm=T)
medianSteps <- median(dailySteps$steps,na.rm=T)
```

Print the mean and median daily steps  

```{r}
print(paste("The Mean number of daily steps is",round(meanSteps,2)))
print(paste("The median number of daily steps is", round(medianSteps,2)))
```

## What is the average daily activity pattern?



## Imputing missing values



## Are there differences in activity patterns between weekdays and weekends?
